---
title: "Morphometrics"
author: "Sara Michele Schaal"
date: "6/28/2018"
output:
  pdf_document: default
  html_document: default
---

The following analyses are done to determine the spawning group for samples of unknown origin. I begin by creating a Linear Discriminate Analysis model for a dataset of individuals with known spawning origin. Using that model I then apply it to my dataset of unknown spawning origin to get posterior probabilities of assignment to each group. 

```{r}
####################################
###### Data & Package Loading ######
####################################

  #install.packages("MASS")
  library(MASS)

  # Load unknown data
  unknown.morph <- read.csv("../data/Morphometrics/morphPlusMetadata20180628.csv")
  
  # Load known data
  known.morph <- read.csv("../data/Morphometrics/GrahamData.csv")
  

###########################
#### Data Manipulation ####
###########################
  
  # convert unknown data to mm
  unknown.morph[,4:26] <- 25.4*unknown.morph[,4:26] 
  
  # add column with red or olive for unknown samples
  unknown.morph$ColorType <- NA
  for(i in 1:nrow(unknown.morph)){
    if(unknown.morph$RedGreenRatio[i] >= 1.3){
      unknown.morph$ColorType[i] <- "Red"
    } else {
      unknown.morph$ColorType[i] <- "Olive"
    }
  }
  
  # convert to cm and take out only the natural log transformed morph data
  unknown.morphlog <- log(unknown.morph[,4:26])  
  known.morphlog <- log(known.morph[,c(16:38)]) # already in mm
  #colnames(known.morphlog)[23] <- "TotalLength"

  # remove data from landmark 11 - excludes gut fullness from affecting results
  unknown.morphlog <- unknown.morphlog[,c("TotalLength", "D1", "D2", "D3", "D4", "D5", 
                                          "D6", "D7", "D8", "D10", "D13", "D14", "D16", 
                                          "D17", "D18", "D19", "D21", "D22")]
  known.morphlog <- known.morphlog[,c( "D1", "D2", "D3", "D4", "D5", "D6", "D7", "D8", 
                                       "D10", "D13", "D14", "D16", "D17", "D18", "D19", 
                                       "D21", "D22")] 
```
 
# Perform Statistical Analysis on Known Data
```{r}
#################
#### Run PCA ####
#################
  # Principle Components Analysis to Account for Total Length of Individuals
  pca.known <- prcomp(known.morphlog, scale. = TRUE)
  pca.known.noScale <- prcomp(known.morphlog)
  summary(pca.known)
  max(pca.known$x[,1])
  min(pca.known$x[,1])
  # Plot PCA
  plot(pca.known$x[,1], pca.known$x[,2])
  screeplot(pca.known)
  hist(pca.known$x[,1])
  
#############################################
#### Regress PC1 on Log Transformed Data ####
#############################################
  # Because PC1 corresponds to variance in morphology due to total length of the 
  # fish we can regressing PC1 on our data to account for that variance in downstream 
  # analyses
  
  # for loop to regress PC1 onto log transformed data and pull out residuals
  regress.known <- matrix(nrow = nrow(pca.known$x), ncol = ncol(pca.known$x))
  for(i in 1:ncol(pca.known$x)){
    regress.known[,i] <- lm(pca.known$x[,1]~known.morphlog[,i])$residuals
  }
  
  # Create dataframe with Spawning Season and Residuals of Regressed Data for DFA
  df.knowndata <- data.frame(known.morph$Season, regress.known)
  
##############################################
#### Perform Linear Discriminate Analysis ####
##############################################
  # Run DFA
  LDAknown.results <- lda(known.morph.Season ~ ., data = df.knowndata)
  LDAknown.predict <- predict(LDAknown.results)
  str(LDAknown.predict)
  
  # Plot DFA 
  plot.data <- data.frame(LDAknown.predict$x, df.knowndata$known.morph.Season)
  colnames(plot.data)[3] <- "Season"
  plot(plot.data$LD1[plot.data$Season == "Red"], 
       plot.data$LD2[plot.data$Season == "Red"], col = "firebrick", 
       ylim = c(-4,4), xlim = c(-8, 7), pch = 19,
       ylab = "DF2", xlab = "DF1", main = "Linear Discriminate Analysis of Cod Ecotype Morphology")
  points(plot.data$LD1[plot.data$Season == "Spring"], 
         plot.data$LD2[plot.data$Season == "Spring"], col = "goldenrod",
         pch = 17)
  points(plot.data$LD1[plot.data$Season == "Winter"],
         plot.data$LD2[plot.data$Season == "Winter"], col = "cornflowerblue",
         pch = 15)
  legend("topleft", legend = c("Red", "Spring", "Winter"), 
         pch = c(19, 17, 15), cex = 0.75,
         col = c("firebrick", "goldenrod", "cornflowerblue"))
  
  # LD1: D13 D14 D8 D7 
  # LD2: D22 D13 D8 D4
```

\newpage
#Results from SPSS 
###Essentially the same as R just has an inverted sign for DF2 for some reason...haven't figured that out yet
```{r}
SPSS.data <- read.csv("/Users/saraschaal/Desktop/GrahamMorph/GrahamSPSSData.csv")
SPSS.LDA <- read.csv("~/Desktop/GrahamMorph/GrahamDataLDA.csv")
SPSS.Log <- SPSS.data[,grep("Ln", names(SPSS.data))] # Same as R
regress.known.SPSS <- SPSS.data[,grep("RES", names(SPSS.data))] # Same as R when PCA is NOT scaled

head(SPSS.LDA$LDA1)
head(SPSS.LDA$LDA2)
head(LDAknown.predict$x[,1])
head(LDAknown.predict$x[,2]) 
# sign is flipped and numbers are SLIGHTLY different from SPSS
# R and SPSS may calculate the LDA slightly differently, 
# but doesn't have large impacts on the results

plot(SPSS.LDA$LDA1[SPSS.LDA$Season == "Spring"], 
     SPSS.LDA$LDA2[SPSS.LDA$Season == "Spring"], col = "darkolivegreen4",
     pch = 17, xlim = c(-8, 5), ylim = c(-3.1, 4),
     xlab = "DF1", ylab = "DF2", main = "SPSS")
points(SPSS.LDA$LDA1[SPSS.LDA$Season == "Winter"], 
     SPSS.LDA$LDA2[SPSS.LDA$Season == "Winter"], col = "cornflowerblue",
     pch = 19)
points(SPSS.LDA$LDA1[SPSS.LDA$Season == "Red"], 
     SPSS.LDA$LDA2[SPSS.LDA$Season == "Red"], col = "firebrick",
     pch = 15)
legend("topleft", legend = c("Red", "Spring", "Winter"), 
         pch = c(15, 17, 19), cex = 0.75,
         col = c("firebrick", "darkolivegreen4", "cornflowerblue"))
```

\newpage
# Remove Red cod and reanalyze
The red cod in the known dataset did not have any length measurements so I had to exclude total length from that analysis. Here I removed red cod from both the known dataset and the unknown dataset to specifiy just spring or winter spawners in my data. This results in a single LD axis.
```{r}
################################################
#### Run LDA for only Spring and Winter cod ####
################################################
  library(randomForest)
  # remove Red cod data 
  known.noRed <- known.morph[known.morph$Season != "Red", ]
  unknown.noRed <- unknown.morph[unknown.morph$ColorType != "Red", ]
  unknown.lognoRed <- log(unknown.noRed[,c("TotalLength", "D1", "D2", "D3", "D4", "D5", 
                                           "D6", "D7", "D8", "D10", "D13", "D14", "D16", 
                                           "D17", "D18", "D19", "D21", "D22")])
  colnames(unknown.lognoRed)[1] <- "D23" 
  unknown.lognoRed <- unknown.lognoRed[,c(2:18, 1)]
  known.noRed$Season <- factor(known.noRed$Season)
  known.lognoRed <- log(known.noRed[,c(16:23, 25, 28:29, 31:34, 36:38)])
  
# Run a Principle Components Analysis on both datasets
  pca.known.noRed <- prcomp(known.lognoRed, scale. = TRUE)
  summary(pca.known.noRed)
  pca.unknown.noRed<- prcomp(unknown.lognoRed, scale. = TRUE)
  
# Plot PCA 
  plot(pca.known.noRed$x[,1], pca.known.noRed$x[,2])
  screeplot(pca.known.noRed)
  hist(pca.known.noRed$x[,1])
  
# for loop to regress PC1 onto log transformed data and pull out residuals
  #Grahams Data
  regress.known.noRed <- matrix(nrow = nrow(pca.known.noRed$x), 
                               ncol = ncol(pca.known.noRed$x))
  for(i in 1:ncol(pca.known.noRed$x)){
    regress.known.noRed[,i] <- lm(pca.known.noRed$x[,1]~known.lognoRed[,i])$residuals
  }
  
  #Sara Data
  regress.unknown.noRed <- matrix(nrow = nrow(pca.unknown.noRed$x), 
                                  ncol = ncol(pca.unknown.noRed$x))
  for(i in 1:ncol(pca.unknown.noRed$x)){
    regress.unknown.noRed[,i] <- lm(pca.unknown.noRed$x[,1]~unknown.lognoRed[,i])$residuals
  }
  
# Create dataframe with data for Linear Discriminate Function Analysis
  df.known.noRed <- data.frame(known.noRed$Season, regress.known.noRed)
  
# Run DFA
  LDA.known.noRed <- lda(known.noRed.Season ~ ., data = df.known.noRed)
  colnames(regress.unknown.noRed) <- c("X1", "X2", "X3", "X4", "X5", "X6", 
                                       "X7", "X8", "X9", "X10", "X11", "X12", 
                                       "X13", "X14", "X15", "X16", "X17", "X18")
  regress.unknown.noRed <- as.data.frame(regress.unknown.noRed)
  LDApredict.known.noRed <- predict(LDA.known.noRed)
  plot(LDA.known.noRed)
  str(LDApredict.known.noRed)
  
  # Evaluate Performance of LDA on Known Data
  df.prediction <- as.data.frame(cbind(as.character(df.known.noRed$known.noRed.Season),
                                       as.character(LDApredict.known.noRed$class),
                                       LDApredict.known.noRed$posterior, LDApredict.known.noRed$x))
  colnames(df.prediction)[1:2] <- c("Known", "Predicted")
  
  incorrect <- NULL
  for(i in 1:nrow(df.prediction)){
    if(df.prediction[i,1] != df.prediction[i,2]){
      incorrect <- as.data.frame(rbind(incorrect, df.prediction[i,]))
    }
  }
  incorrect
  (nrow(regress.known.noRed) - nrow(incorrect))/nrow(regress.known.noRed)*100
  
  # Run DFA on my samples to determine which spawning groups they belong to
  LDApredictnoRed <- predict(LDA.known.noRed, newdata = regress.unknown.noRed)
  
  #Create dataframe for plotting
  df.finalunknown <- cbind(regress.unknown.noRed, LDApredictnoRed$posterior, LDApredictnoRed$x)
  
  # Based on posterior probabilities determine which spawning group samples are from
  df.finalunknown$Spawning <- NA
  for(i in 1:nrow(df.finalunknown)){
    if(df.finalunknown$Spring[i] > 0.5){
     df.finalunknown$Spawning[i] <- "Spring"
    } else {
     df.finalunknown$Spawning[i] <- "Winter"
    }
  }
  
  df.finalunknown <- data.frame(df.finalunknown, unknown.noRed$Sample)
  colnames(df.finalunknown)[19:23] <- c("LDAMorphSpring", "LDAMorphWinter", "LD1Morph","LDAMorphAssign","SampleID")
  write.csv(df.finalunknown, "../data/Morphometrics/SamplingSpawningAssignment.csv")
  
  
  colnames(df.known.noRed)[1] <- "SEASON"
  #Subset data into training and validation datasets to test how well it works on known data
  trainMorph <- sample(nrow(df.known.noRed), 0.7*nrow(df.known.noRed), replace = FALSE)
  TrainMorphData <- df.known.noRed[trainMorph,]
  ValidMorphData <- df.known.noRed[-trainMorph,]
 
  #Create RFmodel on Training Data
  RFMorphModel <- randomForest(SEASON ~ ., data = TrainMorphData, importance = TRUE)
  
  #Run RFmodel on the Training Dataset
  predTrainMorph <- predict(RFMorphModel, TrainMorphData, type = "class")
  table(predTrainMorph, TrainMorphData$SEASON)
  
  #Run RFmodel on the Validation Dataset
  predValidMorph <- predict(RFMorphModel, ValidMorphData, type = "class")
  table(predValidMorph, ValidMorphData$SEASON)
  # 86% accuracy on validation dataset
  
  # Run Model from Training Data on full dataset
  RFAssignMorph <- predict(RFMorphModel, df.known.noRed, type = "class")
  RFMorphPred <- predict(RFMorphModel, df.known.noRed, type = "prob")
  table(RFAssignMorph, df.known.noRed$SEASON)
  # 96% on full dataset
  
  df.RFMorphGraham <- as.data.frame(cbind(df.known.noRed, RFMorphPred, RFAssignMorph))
  
  # Check accuracy of the models
  RFMorphcorrect <- NULL
  RFMorphincorrect <- NULL
  for(i in 1:nrow(df.RFMorphGraham)){
    if(df.RFMorphGraham$SEASON[i] == df.RFMorphGraham$RFAssignMorph[i]){
      RFMorphcorrect <- c(RFMorphcorrect, as.character(df.RFMorphGraham$SEASON[i]))
    } else {
      RFMorphincorrect <- c(RFMorphincorrect, as.character(df.RFMorphGraham$SEASON[i]))
    }
  }
  100-(length(RFMorphincorrect)/length(RFMorphcorrect)*100)
  
 # Use model to predict unknown samples
  RFMorphSara <- predict(RFMorphModel, regress.unknown.noRed, type = "prob")
  colnames(RFMorphSara) <- c("RFMorphSpring", "RFMorphWinter")
  RFMorphAssignSara <- as.character(predict(RFMorphModel, regress.unknown.noRed))
  
  df.RFMorphSara <- as.data.frame(cbind(RFMorphSara, RFMorphAssignSara, unknown.noRed,
                                        df.finalunknown[19:23]))
  
  write.csv(df.RFMorphSara, "../data/Morphometrics/MorphAnalyses.csv")
  
## Plot
  plot(df.RFMorphSara$RFMorphSpring[df.RFMorphSara$RFMorphSpring > 0.5], 
       df.RFMorphSara$RFMorphWinter[df.RFMorphSara$RFMorphWinter < 0.5], 
       col = "darkolivegreen4", pch = 19, ylim = c(0,1), xlim = c(0,1),
       xlab = "Probability of Spring Spawner", ylab = "Probability of Winter Spawner",
       main = "Random Forest Posterior Probabilities")
  points(df.RFMorphSara$RFMorphSpring[df.RFMorphSara$RFMorphSpring < 0.5], 
       df.RFMorphSara$RFMorphWinter[df.RFMorphSara$RFMorphWinter > 0.5], 
       col = "cornflowerblue", pch = 19, ylim = c(0,1), xlim = c(0,1),
       xlab = "Probability of Spring Spawner", ylab = "Probability of Winter Spawner")
  legend("topright", legend = c("Winter", "Spring"), 
         col = c("cornflowerblue", "darkolivegreen4"), pch = 19)
  
  hist(df.RFMorphSara$RFMorphSpring[df.RFMorphSara$RFMorphSpring > 0.5], breaks = 10, 
       col = "darkolivegreen4", xlim = c(0,1), ylim = c(0, 20), 
       xlab = "Probability of Spring Spawner", 
       main = "RF Probability of Assignment to Spring Spawning Cod")
  hist(df.RFMorphSara$RFMorphSpring[df.RFMorphSara$RFMorphSpring < 0.5], breaks = 10, 
       col = "cornflowerblue", add = TRUE)
  legend("topright", legend = c("Winter", "Spring"), 
         fill = c("cornflowerblue", "darkolivegreen4"))
  
  
  pdf(file = "../figures/PosteriorProbabilityOfAssignment.pdf")
  par(mfrow = c(1,1))
  plot(df.finalunknown$LDAMorphSpring[df.finalunknown$LDAMorphSpring > 0.5], 
       df.finalunknown$LDAMorphWinter[df.finalunknown$LDAMorphWinter < 0.5], 
       col = "darkolivegreen4", pch = 19, ylim = c(0, 1), xlim = c(0,1),
       xlab = "Probability of Spring Spawner", ylab = "Probability of Winter Spawner",
       main = "LDA Posterior Probabilities")
  points(df.finalunknown$LDAMorphSpring[df.finalunknown$LDAMorphSpring < 0.5], 
         df.finalunknown$LDAMorphWinter[df.finalunknown$LDAMorphWinter > 0.5], 
         col = "cornflowerblue", pch = 19)
  legend("topright", legend = c("Winter", "Spring"), 
         col = c("cornflowerblue", "darkolivegreen4"), pch = 19)
  #dev.off()
  
pdf(file = "../figures/SpawningAssignment.pdf")

  hist(df.finalunknown$LDAMorphSpring[df.finalunknown$LDAMorphSpring > 0.5], breaks = 10, 
       col = "darkolivegreen4", xlim = c(0,1), ylim = c(0, 20), 
       xlab = "Probability of Spring Spawner", 
       main = "LDA Probability of Assignment to Spring Spawning Cod")
  hist(df.finalunknown$LDAMorphSpring[df.finalunknown$LDAMorphSpring < 0.5], breaks = 10, 
       col = "cornflowerblue", add = TRUE)
  legend("topright", legend = c("Winter", "Spring"), 
         fill = c("cornflowerblue", "darkolivegreen4"))
  #dev.off()
  
```

